---
title: "Education data cleaning"
author: "Stephanie Veazie"
date: "2022-12-02"
output:
  word_document: default
  html_document: default
---

## Download data 

1. Go to https://data.census.gov/cedsci/

2. Select "advanced search." 

3. Using the filter pane on the left-hand side of the page, under "Geography", select "State" then "All states with in the US, Puerto Rico, and the Island Areas." 

4. Under "Topics," select "Education" then "Educational attainment."

5. Under "Topics" select "Race and ethnicity" then "Race and Ethnicity" again.

6. Under "Surveys" select "American Community Survey", then "5-Year Estimates" then "Detailed Tables" 

7. Click blue "Search" button in the bottom right-hand corner.

8. Select "C15002H" for White Alone, Not Hispanic or Latino. 

9. Under drop-down menu, select "2019: ACS 5-year estimates subject tables." Download as a csv file. 

10. In the results pane, select "C15002B" for Black Alone. 

11. Select "2019: ACS 5-year estimates subject tables." Download as a csv file

12. Repeat steps 8-11 but for 2015 ACS 5-year estimates. 

## Load libraries

```{r}
# Load libraries
library(tidyverse)
```

## Import files

```{r}
# Import files
white_edu_2019 <- read_csv("../../data/raw/us-census/ACSDT5Y2019.C15002H-2024-08-14T232424.csv")
black_edu_2019 <- read_csv("../../data/raw/us-census/ACSDT5Y2019.C15002B-2024-08-14T232533.csv")
```

## Create function to clean csv file

```{r}
clean_edu_census_file <- function(file_name){
  
edu <- file_name

# Select only label and estimate columns 
edu_2 <- edu %>% select(contains("Label") | contains("!!Estimate"))

# Create a gender vector
gender_vec <- c("both_genders", rep("male", 5), rep("female", 5))

# Combine gender vector + dataset 
edu_3 <- cbind(gender_vec, edu_2)

# Rename label col
edu_4 <- edu_3 %>% rename("variable" = "Label (Grouping)")

# Create new col with cleaner variable names
edu_5 <- edu_4 %>% mutate(variable_renamed =
                                        case_when(
                                          grepl("Total:", variable) ~ "total", 
                                          grepl("Male:", variable) ~ "total", 
                                          grepl("Female:", variable) ~ "total", 
                                          grepl("Less than high school diploma", variable) ~ "less_than_hs", 
                                          grepl("High school graduate", variable) ~ "hs_grad", 
                                          grepl("Some college or associate's degree", variable) ~ "some_college", 
                                          grepl("Bachelor's degree or higher", variable) ~ "college_degree", 
                                        ))

# Rearrange and check results
edu_6 <- edu_5[,c(1,55,2:54)]

# Remove !!Estimate from col names
edu_7 <- edu_6 %>% rename_with(~ gsub("!!Estimate", "", .))

# Combine variable and gender cols
edu_7$variable_final <- paste(edu_7$variable_renamed, edu_7$gender_vec, sep = "_")

# Rearrange cols
edu_8 <- edu_7[,c(56,1:55)]

# Drop unneeded cols
edu_9 <- edu_8 %>% subset(select = -c(gender_vec, variable_renamed, variable))

# Transpose data
edu_10 <- as.data.frame(t(edu_9))

# Rename columns
colnames(edu_10) <- edu_10[1,]

# Delete duplicate first row
edu_11 <- edu_10[-1,]

# Add new column of state names
edu_11$state <- rownames(edu_11)

# Rearrange
edu_12 <- edu_11[,c(12,1:11)]

# Delete row names
rownames(edu_12) <- NULL

# Transform rows 2:12 to numeric
edu_12[,2:12] <- lapply(edu_12[,2:12], as.numeric)

return(edu_12)
}
```

# Apply function to get cleaned datasets

```{r}
# Create cleaned dataset
white_edu_clean_2019 <- clean_edu_census_file(white_edu_2019)
black_edu_clean_2019 <- clean_edu_census_file(black_edu_2019)

# Add race to col names 2-12
colnames(white_edu_clean_2019)[2:12] <- paste0(colnames(white_edu_clean_2019)[2:12], "_white")
colnames(black_edu_clean_2019)[2:12] <- paste0(colnames(black_edu_clean_2019)[2:12], "_black")
```

# Combine datasets

```{r}
# Combine datasets
comb_edu_2019 <- left_join(white_edu_clean_2019, black_edu_clean_2019, by = "state")

# Calculate % of white & black pop w/out college degree & B-W RR
comb_edu_2019_2 <- comb_edu_2019 %>%
  mutate(
    percent_no_college_white = 1 -
      ((college_degree_male_white + college_degree_female_white) / total_both_genders_white), 
    
    percent_no_college_black = 1 - 
      ((college_degree_male_black + college_degree_female_black) / total_both_genders_black),
    
    edu_RR = percent_no_college_black / percent_no_college_white
  ) 

comb_edu_2019_2
```

# Export

```{r}
# Add year
comb_edu_2019_final <- comb_edu_2019_2 %>% mutate(year = 2019)

# Rearrange cols
comb_edu_2019_final <- comb_edu_2019_final[,c(1,27,2:26)]

# Export
write_csv(comb_edu_2019_final, "../../data/cleaned/edu-attainment-clean.csv")
```

