---
title: "Employment data cleaning"
author: "Stephanie Veazie"
date: "2022-12-2"
output:
  word_document: default
---

## Download data

1. Go to https://data.census.gov/cedsci/

2. Select "advanced search." 

3. On the filter on the left-hand side of the page, under "geographies" select "state" then select "all states with in the US, Puerto Rico, and the Island Areas." 

4. Under "topics," select "employment" then "employment and labor force status."

5. Click "search" 

6. Select "S2301 Employment status"

7. Under drop-down menu, select "2019: ACS 5-year estimates subject tables"

8. Select "transpose."

9. Download as a csv file. 

10. Repeat steps 7-9 for "2014: ACS 5-year estimates subject tables"

# Load libraries

```{r}
# Load libraries
library(tidyverse)
library(magrittr)
```

# Import files

```{r}
employ_2019 <- read_csv("../../data/raw/us-census/ACSST5Y2019.S2301-2024-08-15T201852.csv")
```

# Create function to clean 2019 csv file

```{r}
clean_employ_2019_census_file <- function(file_name){

# Create dataset of columns that report variable name and data by race
selected_employment_data <- file_name %>% 
  select(`Label (Grouping)`, 
         `Population 16 years and over`, 
         starts_with("Population 16 years and over!!R") ) %>%
  select(-`Population 16 years and over!!RACE AND HISPANIC OR LATINO ORIGIN`)

names(selected_employment_data)
# Simplify the column names
names(selected_employment_data)<-c("variable", 
                                   "total",
                                   "white", 
                                   "black", 
                                   "AIAN", 
                                   "asian", 
                                   "NHPI", 
                                   "other_race", 
                                   "two_races", 
                                   "hispanic", 
                                   "white_non_hispanic")

# Confirm change
names(selected_employment_data)

state_indices <- seq(1, 664, by = 13)

selected_employment_data %<>% 
  mutate(state_vector = case_when(row_number() %in% state_indices ~ variable,
                           !(row_number() %in% state_indices) ~ NA_character_)
  )

selected_employment_data %<>% mutate(grouping_indicator = ifelse(is.na(state_vector), 0, 1),
                                     group_sum = cumsum(grouping_indicator))

states_dat <- selected_employment_data %>% group_by(group_sum) %>% summarise(state_vector = first(state_vector))
selected_employment_data2 <- merge(states_dat, 
                                   selected_employment_data, 
                                   by = "group_sum",
                                   all.y = T)

selected_employment_data2 %<>%
  select(-state_vector.y, -starts_with("group")) %>% rename(state_vector = state_vector.x)

#select the total population estimate, labor force participation and the unemployment estimates only.

selected_employment_data2 %<>% mutate(estimate_ind = ifelse(str_detect(variable, "Estimate"), 1, 0))

selected_employment_data2 %<>% group_by(state_vector) %>% mutate(est_count = cumsum(estimate_ind))

selected_employment_data3 <- selected_employment_data2 %>% filter(estimate_ind == 1 & est_count %in% c(1, 2, 4)) %>% 
  # added 2 above to pull the labor participation rate
  mutate(variable = 
           case_when( # changed this from if_else to case_when since there are now 3 categories
             estimate_ind == 1 & est_count == 1 ~ "total_pop",
             estimate_ind == 1 & est_count == 2 ~ "labor_force", 
             estimate_ind == 1 & est_count == 4 ~ "unemployment"))

# Remove ',' and '%' from data in order to create numerical data
selected_employment_data4 <- selected_employment_data3 %>% mutate_all(~ str_replace(., "%", ""))
selected_employment_data4 <- selected_employment_data4 %>% mutate_all(~ str_replace_all(., ",", ""))

selected_employment_data4 %<>% select(-estimate_ind, -est_count)

# Pivot wider so each state's data is on 1 row.
tidy_unemployment <- selected_employment_data4 %>% 
  pivot_wider(names_from= variable, 
              values_from = c("total", "white", "black", "AIAN", "asian", "NHPI", "other_race", "two_races", "hispanic", "white_non_hispanic"),
              names_glue="{variable}_{.value}")

# Subset data
tidy_unemployment_black_white <- tidy_unemployment %>% select(state_vector, ends_with(c("white_non_hispanic", "black")))

# Create RR
tidy_unemployment_black_white %<>%  
  mutate(total_pop_white_non_hispanic = as.numeric(total_pop_white_non_hispanic),
         labor_force_white_non_hispanic = as.numeric(labor_force_white_non_hispanic)/100, 
         unemployment_white_non_hispanic = as.numeric(unemployment_white_non_hispanic)/100,
         total_pop_black = as.numeric(total_pop_black),
         labor_force_black = as.numeric(labor_force_black)/100,
         unemployment_black = as.numeric(unemployment_black)/100)

# Add column with White:Black ratio and difference in unemployment
tidy_unemployment_bw <- tidy_unemployment_black_white %>%
  mutate( 
    bw_unemploy_RR = unemployment_black / unemployment_white_non_hispanic)

## Calculate CI for unemployment RR

return(tidy_unemployment_bw)
}
```

# Apply function to get cleaned datasets

```{r}
employ_2019_clean <- clean_employ_2019_census_file(employ_2019)
```

# Export

```{r}
# Add year
comb_employ_2019_final <- employ_2019_clean %>% mutate(year = 2019)

# Rearrange cols
comb_employ_2019_final <- comb_employ_2019_final[,c(1,9,2:8)]

comb_employ_2019_final <- comb_employ_2019_final %>% rename(
  state = state_vector
)

# Export
write_csv(comb_employ_2019_final, "../../data/cleaned/unemployment-clean.csv")
```
