---
title: "Income data cleaning"
author: "Stephanie Veazie"
date: "2022-12-05"
output:
  word_document: default
  html_document: default
---

## Download data

1. Go to https://data.census.gov/cedsci/

2. Select "Advanced search." 

3. Using the filter pane on the left-hand side of the page, under "Geography", select "State" then "All states with in the US, Puerto Rico, and the Island Areas." 

4. Under "Topics," select "Income and poverty" then "Income and earnings" then "Income and earnings"

5. Click blue "Search" button in the bottom right-hand corner.

6. In the top search bar, enter in "S1903" then select "S1903 Median income in the past 12 months (in 2019 inflation-adjusted dollars)" from the menu that appears. 

7. Under drop-down menu, select "2019: ACS 5-year estimates subject tables"

8. Click "transpose." 

9. Download as a csv file. 

10. Repeat steps #6-9 for 2014 5-year data (in 2014 inflation-adjusted dollars)

## Load libraries

```{r}
# Load libraries
library(tidyverse)
```

## Import files

```{r}
income_2019 <- read_csv("../../data/raw/us-census/ACSST5Y2019.S1903-2024-08-15T225701.csv")
```

# Create function to clean 2019 data

```{r}
clean_income_2019_census_file <- function(file_name){

median_income <- income_2019[, 1:13]

names(median_income)

names(median_income) <- c("variable", 
                           "blank_column", 
                           "total_all_categories", 
                           "second_blank_column",
                           "white", 
                           "black", 
                           "american_indian",
                           "asian",
                           "native_hawaiian",
                           "other_race", 
                           "two_races", 
                           "hispanic",
                           "white_non_hispanic")

state_indices <- seq(1, 520, by = 10) 

median_income %<>% 
  mutate(state_vector = case_when(row_number() %in% state_indices ~ variable,
                           !(row_number() %in% state_indices) ~ NA_character_)
  )

median_income %<>% mutate(grouping_indicator = ifelse(is.na(state_vector), 0, 1),
                                     group_sum = cumsum(grouping_indicator))

states_dat <- median_income %>% group_by(group_sum) %>% summarise(state_vector = first(state_vector))
median_income2 <- merge(states_dat, 
                                   median_income, 
                                   by = "group_sum",
                                   all.y = T)

median_income2 %<>%
  select(-state_vector.y, -starts_with("group"), -blank_column, -second_blank_column) %>% 
  rename(state_vector = state_vector.x)

#select the total population estimate and the median only.

median_income2 %<>% mutate(estimate_ind = ifelse(str_detect(variable, "Estimate"), 1, 0))

median_income2 %<>% group_by(state_vector) %>% mutate(est_count = cumsum(estimate_ind))

median_income3 <- median_income2 %>% filter(estimate_ind == 1 & est_count %in% c(1, 3)) %>% 
  mutate(variable = ifelse(estimate_ind == 1 & est_count == 1, "number", "median_income"))

# Remove ',' from data in order to create numerical data
median_income3 <- median_income3 %>% mutate_all(~ str_replace_all(., ",", ""))

# Limit to just median income rows
median_income4 <- median_income3 %>% filter(variable == "median_income")

median_income4 %<>% select(-estimate_ind, -est_count)

# Filter to race/ethnic groups of interest
median_income5 <- median_income4 %>% select(state_vector, variable, black, white_non_hispanic)

# Pivot wider so each state's data is on 1 row.
tidy_income <- median_income5 %>% 
  pivot_wider(names_from= variable, 
              values_from = c("black", "white_non_hispanic"),
              names_glue="{variable}_{.value}")

# Save as numeric data
tidy_income %<>%  
  mutate(median_income_black = as.numeric(median_income_black),
         median_income_white_non_hispanic = as.numeric(median_income_white_non_hispanic))

# Create new dataset w/ RRs
tidy_income %<>%  mutate(wb_income_RR = median_income_white_non_hispanic / median_income_black)

return(tidy_income)
}
```

# Apply functions

```{r}
# Create cleaned dataset
income_clean_2019 <- clean_income_2019_census_file(income_2019)
```

# Export

```{r}
# Add year
income_2019_final <- income_clean_2019 %>% mutate(year = 2019)

# Rename state col
income_2019_final  <- income_2019_final  %>% rename(state = state_vector)

# Rearrange cols
income_2019_final  <- income_2019_final[,c(1,5,2:4)]

# Export
write_csv(income_2019_final , "../../data/cleaned/income-clean.csv")
```
