---
title: "Incarceration data cleaning with new denominators"
author: "Stephanie Veazie"
date: "2023-05-05"
output: html_document
---

# Import data from Bureau of Justice Statistics

1. Go to https://csat.bjs.ojp.gov/advanced-query

2. Under "Query type" select "offender characteristics." 

3. Under "category" select "year-end population."

4. Under "sex" select "all."

5. Under "variable 1" select "race/hispanic ethnicity of inmate." 

6. Under "jurisdiction" select "all."

7. Under "year" select 2010-2019. 

8. Hit "submit." 

Warning from BJS website: Users are cautioned against making state-to-state comparisons with CSAT-Prisoners because states have very different offender profiles. Some states have unified prison and jail systems, while in other states only offenders with a sentence of more than one year serve sentences in prison. NCRP only includes offenders serving sentences in prisons.

9. Select "ok. 

10. Select "export to Excel." 

11. Rename file "bjs-incarceration-by-race-state-2010-2019-aug-22-24-dl.xlsx

# Import data from US Census Bureau website

1. Go to the US Census Bureau's Population & Housing Unit Estimates website-
https://www.census.gov/programs-surveys/popest/technical-documentation/research/evaluation-estimates/2020-evaluation-estimates/2010s-state-detail.html

2. Download the Excel sheet titled "Annual State Resident Population Estimates for 6 Race Groups (5 Race Alone Groups and one multiple race group) by Age, Sex, and Hispanic Origin: April 1, 2010 to July 1, 2019; April 1, 2020; and July 1, 2020 (SC-EST2020-ALLDATA6) [<1.0 MB]" 

3. Download the accompanying "File Layout" PDF 

# Load Census Bureau data into R 

```{r}
# Load libraries
library(tidyverse)
library(dplyr)
library(readxl)
library(magrittr)

# Import CSV
residents<- read_csv("../../data/raw/us-census/SC-EST2020-ALLDATA6.csv")
```

```{r}
# Select variables of interest
selected_residents<- select(residents, c("STATE", "NAME", "SEX", "ORIGIN", "RACE", "AGE", 
                                         "POPESTIMATE2010", 
                                         "POPESTIMATE2011", 
                                         "POPESTIMATE2012", 
                                         "POPESTIMATE2013", 
                                         "POPESTIMATE2014",
                                         "POPESTIMATE2015", 
                                         "POPESTIMATE2016",
                                         "POPESTIMATE2017", 
                                         "POPESTIMATE2018",
                                         "POPESTIMATE2019"))


# Filter to non-hispanic individuals
selected_residents<- filter(selected_residents, ORIGIN == 1)
```

```{r}
# Filter to all sexes
selected_residents %<>% filter(SEX == 0)
```

```{r}
# Filter to Black or White individuals
selected_residents<- filter(selected_residents, RACE == 1 | RACE == 2) 
```

```{r}
# Filter to those aged 15-64
selected_residents <- selected_residents %>% filter(AGE %in% (15:64))
```

```{r}
# Group by state & race
by_state_race<- group_by(selected_residents, NAME, RACE)

# Sum residential info by group
resident_summary<- as_data_frame(
  summarize(by_state_race, 
            sum_2010 = sum(POPESTIMATE2010),
            sum_2011 = sum(POPESTIMATE2011),
            sum_2012 = sum(POPESTIMATE2012),
            sum_2013 = sum(POPESTIMATE2013),
            sum_2014 = sum(POPESTIMATE2014),
            sum_2015 = sum(POPESTIMATE2015), 
            sum_2016 = sum(POPESTIMATE2016), 
            sum_2017 = sum(POPESTIMATE2017), 
            sum_2018 = sum(POPESTIMATE2018), 
            sum_2019 = sum(POPESTIMATE2019))
)
```

```{r}
# Label race data
resident_summary <- resident_summary %>%
  mutate(
    race = case_when(
      RACE == 1 ~ "white_nh", 
      RACE == 2 ~ "black_nh")) %>% 
  select(-RACE)

# pivot longer
resident_summary2 <- resident_summary %>% 
  pivot_longer(! c(NAME, race), names_to = "year", values_to = "num_residents")

resident_summary2 %<>% mutate(year = str_sub(year, -4, -1))

resident_summary2 %<>% rename(state_vector = NAME)
rm(resident_summary)

```

# Load BJS data into R

```{r}
incarceration<- read_xlsx("../../data/raw/csat-bjs/bjs-incarceration-by-race-state-2010-2019-aug-22-24-dl.xlsx", range = "A2:L253", na = c("*", "NA"))
```

```{r}
# Rename columns
colnames(incarceration)<- c("state", "race_ethnicity_of_inmate", 
                            "year_2010", "year_2011", "year_2012", "year_2013", "year_2014",
                            "year_2015", "year_2016", "year_2017", "year_2018", "year_2019")

state_indices <- seq(1, 251, by = 5)

#does not contain Puerto Rico and DC at end

incarceration %<>% 
  mutate(state_vector = case_when(row_number() %in% state_indices ~ state,
                           !(row_number() %in% state_indices) ~ NA_character_)
  )

incarceration %<>% mutate(grouping_indicator = ifelse(is.na(state_vector), 0, 1),
                                     group_sum = cumsum(grouping_indicator))

states_dat <- incarceration %>% group_by(group_sum) %>% summarise(state_vector = first(state_vector))
incarceration2 <- merge(states_dat, 
                        incarceration, 
                        by = "group_sum",
                        all.y = T)

incarceration2 %<>%
  select(-state_vector.y, -state, -starts_with("group")) %>% rename(state_vector = state_vector.x)

capwords <- function(s, strict = FALSE) {
    cap <- function(s) paste(toupper(substring(s, 1, 1)),
                  {s <- substring(s, 2); if(strict) tolower(s) else s},
                             sep = "", collapse = " " )
    sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
}

incarceration2 %<>% mutate(state_vector = capwords(tolower(state_vector)))

incarceration2 %<>% filter(race_ethnicity_of_inmate %in% c("Black, non-Hispanic", "White, non-Hispanic"))

incarceration2 %<>% mutate(race_ethnicity_of_inmate = 
                             case_when(race_ethnicity_of_inmate == "Black, non-Hispanic" ~ "black_nh", 
                                       race_ethnicity_of_inmate == "White, non-Hispanic" ~ "white_nh"))

incarceration2 %<>% rename(num_incarcerated_2010 = year_2010,
                           num_incarcerated_2011 = year_2011,
                           num_incarcerated_2012 = year_2012,
                           num_incarcerated_2013 = year_2013,
                           num_incarcerated_2014 = year_2014,
  
                           num_incarcerated_2015 = year_2015,
                           num_incarcerated_2016 = year_2016,    
                           num_incarcerated_2017 = year_2017,
                           num_incarcerated_2018 = year_2018,
                           num_incarcerated_2019 = year_2019)
rm(incarceration)
rm(states_dat)

# pivot longer
incarceration3 <- incarceration2 %>% pivot_longer(! c(state_vector, race_ethnicity_of_inmate), names_to = "year", values_to = "num_incarcerated")

incarceration3 %<>% mutate(year = str_sub(year, -4, -1))

incarceration3 %<>% rename(race = race_ethnicity_of_inmate)

```

# Combine BJS & US Census Bureau Datasets

```{r}
setdiff(incarceration3$state_vector, resident_summary2$state_vector)
setdiff(resident_summary2$state_vector, incarceration3$state_vector) #No DC in incarceration data

tidy_incarceration <- merge(resident_summary2, incarceration3, by = c("state_vector", "race", "year"), all.x = T)

tidy_incarceration <- tidy_incarceration %>% mutate(incarceration_rate_per100k = (num_incarcerated/num_residents)*100000)
```

# Export data for viz purposes

```{r}
# Find ave incarceration rate
tidy_incarceration2 <- tidy_incarceration %>% select(-num_residents, -num_incarcerated)

# Pivot wider
tidy_incarceration_wide <- tidy_incarceration2 %>% pivot_wider(names_from = c(year, race), values_from = incarceration_rate_per100k)

tidy_incarceration_wide %<>% rename_with(~ paste0("incarceration_rate_", .))

# Find 5-year averages 

tidy_incarceration_wide %<>%
  mutate(white_incarceration_rate_2010_2014 = 
           rowMeans(tidy_incarceration_wide[,12:16], na.rm = TRUE), 
         
         white_incarceration_rate_2015_2019 = 
           rowMeans(tidy_incarceration_wide[,17:21], na.rm = TRUE), 
         
         black_incarceration_rate_2010_2014 = 
           rowMeans(tidy_incarceration_wide[,2:6], na.rm = TRUE), 
         
         black_incarceration_rate_2015_2019 = 
           rowMeans(tidy_incarceration_wide[,7:11], na.rm = TRUE)
  )

orig_incarc <- tidy_incarceration_wide %>% mutate(
  bw_incarc_rr_2019 = black_incarceration_rate_2015_2019 / white_incarceration_rate_2015_2019, 
  bw_incarc_rr_2014 = black_incarceration_rate_2010_2014 / white_incarceration_rate_2010_2014
)

write_csv(orig_incarc, "../../data/cleaned/incarceration-ncrp-clean-for-viz.csv")
```

```{r}
# 2019
summary_stats_2019 <- tidy_incarceration %>% 
  group_by(state_vector, race) %>% 
  summarise(ave_incarceration_rate_per100k =  mean(incarceration_rate_per100k, na.rm = T),
            num_years_included = sum(!is.na(num_incarcerated)),
            ave_population = mean(num_residents),
            ave_incarcerated = mean(num_incarcerated, na.rm = T))

summary_stats2_2019 <- summary_stats_2019 %>% pivot_wider(names_from = race, values_from = c(ave_incarceration_rate_per100k, num_years_included, ave_population, ave_incarcerated))

summary_stats2_2019 %<>% mutate(bw_rr_incarc = 
                                  ave_incarceration_rate_per100k_black_nh/
                                  ave_incarceration_rate_per100k_white_nh)

incarceration_final_2019 <- summary_stats2_2019
```

```{r}
incarceration_final_2019 <- incarceration_final_2019 %>% mutate(year = 2019)

# Export data
write.csv(incarceration_final_2019,"../../data/cleaned/incarceration-ncrp-clean.csv", row.names = FALSE)
```


