---
title: "4b. Alternative incarceration data cleaning"
author: "Stephanie Veazie"
date: "2024-08-08"
output: html_document
---

```{r, load libraries}
library(tidyverse)
library(magrittr)
```

Load data from National Prisoner Survey (NPS)

https://www.icpsr.umich.edu/web/NACJD/studies/38871

1.	Available from University of Michigan Institute for Social Research 
2.	Requires signing up for an account
3.	Click on “Download” then “R” 
4.	Agree to terms of use
5.	Download as a Zip file

To get the codebook, starting from the link above, click on "Explore data" then "Download codebook."

```{r, NPS data}
# load & save data
load("../../data/raw/nps/ICPSR_38871/DS0001/38871-0001-Data.rda")

nps <- da38871.0001

# select applicable variables
nps_2 <- nps %>% select(
  YEAR, 
  STATEID, 
  STATE, 
  BLACKM, # non-hispanic
  BLACKF, # non-hispanic
  WHITEM, # non-hispanic
  WHITEF  # non-hispanic
)

# Filter to 2010-2019
nps_3 <- nps_2 %>% filter(YEAR %in% c(2015, 2016, 2017, 2018, 2019))

# Combine female & male columns
nps_4 <- nps_3 %>% mutate(
  incarcerated_black_nh = BLACKM + BLACKF, 
  incarcerated_white_nh = WHITEM + WHITEF
)

# Select only the variables of interest
nps_5 <- nps_4 %>% select(
  YEAR, 
  STATE, 
  incarcerated_white_nh, 
  incarcerated_black_nh
)

# Pivot wider so each state's data is on 1 row.
nps_wide <- nps_5 %>% pivot_wider(names_from= "YEAR",  
                                  values_from = c(incarcerated_white_nh, 
                                                  incarcerated_black_nh),
                                  names_glue="{.value}_{YEAR}")
```

# Import data from US Census Bureau website

1. Go to the US Census Bureau's Population & Housing Unit Estimates website-
https://www.census.gov/programs-surveys/popest/technical-documentation/research/evaluation-estimates/2020-evaluation-estimates/2010s-state-detail.html

2. Download the Excel sheet titled "Annual State Resident Population Estimates for 6 Race Groups (5 Race Alone Groups and Two or More Races) by Age, Sex, and Hispanic Origin: April 1, 2010 to July 1, 2019; April 1, 2020; and July 1, 2020 (SC-EST2020-ALLDATA6) [<1.0 MB]" 

3. Download the accompanying "File Layout" PDF 

```{r}
# Import CSV
residents<- read_csv("../../data/raw/us-census/SC-EST2020-ALLDATA6.csv")
```

```{r}
# Select variables of interest, 2010-2019
selected_residents<- select(residents, c("STATE", "NAME", "SEX", "ORIGIN", "RACE", "AGE", 
                                         "POPESTIMATE2015", 
                                         "POPESTIMATE2016",
                                         "POPESTIMATE2017", 
                                         "POPESTIMATE2018",
                                         "POPESTIMATE2019"))

# Filter to non-hispanic individuals
selected_residents_2<- filter(selected_residents, ORIGIN == 1)
```

```{r}
# Filter to all sexes
selected_residents_2 %<>% filter(SEX == 0)
```

```{r}
# Filter to Black or White individuals
selected_residents_3 <- filter(selected_residents_2, RACE == 1 | RACE == 2) 
```

```{r}
# Filter to those aged 15-64
selected_residents_4 <- selected_residents_3 %>% filter(AGE %in% (15:64))
```

```{r}
# Group by state & race
by_state_race<- group_by(selected_residents_4, NAME, RACE)

# Sum residential info by group
resident_summary<- as.data.frame(
  summarize(by_state_race, 
            residents_2015 = sum(POPESTIMATE2015), 
            residents_2016 = sum(POPESTIMATE2016), 
            residents_2017 = sum(POPESTIMATE2017), 
            residents_2018 = sum(POPESTIMATE2018), 
            residents_2019 = sum(POPESTIMATE2019))
)

# Label race data
resident_summary<- resident_summary %>%
  mutate(
    race = case_when(
      RACE == 1 ~ "white_nh", 
      RACE == 2 ~ "black_nh"))

# Remove duplicative race row
resident_summary <- resident_summary[,-2]
```

```{r}
# Pivot wider so each state's data is on 1 row.
resident_summary_wide <- resident_summary %>% pivot_wider(names_from= race, values_from = c(
  "residents_2015", "residents_2016", "residents_2017", "residents_2018", "residents_2019"
  ), names_glue="{.value}_{race}")
```

### Import US Census region & division data

Download file on US Census Regions & Divisions from https://github.com/cphalpert/census-regions/blob/master/us%20census%20bureau%20regions%20and%20divisions.csv

```{r}
# Import data on US Census Bureau region & division
us_census<- read.csv("../../data/raw/us-census/us_census_bureau_regions_and_divisions.csv")

# Remove DC
# us_census<- us_census %>% filter(State !="District of Columbia")

# Merge with existing dataset
tidy_residents <- merge(resident_summary_wide, us_census, by.x = "NAME", by.y = "State")
```

Merge incarceration & residents datasets

```{r}
# merge datasets
incarceration <- merge(tidy_residents, nps_wide, by.x = "State.Code", by.y = "STATE")

# Create new variables
incarceration_2 <- incarceration %>% mutate(
  
          # white
          
          white_incarceration_rate_2015 = incarcerated_white_nh_2015 * 100000/
            residents_2015_white_nh, 
          
          white_incarceration_rate_2016 = incarcerated_white_nh_2016 * 100000/
            residents_2016_white_nh, 
          
          white_incarceration_rate_2017 = incarcerated_white_nh_2017 * 100000/
            residents_2017_white_nh, 
          
          white_incarceration_rate_2018 = incarcerated_white_nh_2018 * 100000/
            residents_2018_white_nh, 
          
          white_incarceration_rate_2019 = incarcerated_white_nh_2019 * 100000/
            residents_2019_white_nh, 
          
          # black

          black_incarceration_rate_2015 = incarcerated_black_nh_2015 * 100000/
            residents_2015_black_nh, 
          
          black_incarceration_rate_2016 = incarcerated_black_nh_2016 * 100000/
            residents_2016_black_nh, 
          
          black_incarceration_rate_2017 = incarcerated_black_nh_2017 * 100000/
            residents_2017_black_nh, 
          
          black_incarceration_rate_2018 = incarcerated_black_nh_2018 * 100000/
            residents_2018_black_nh, 
          
          black_incarceration_rate_2019 = incarcerated_black_nh_2019 * 100000/
            residents_2019_black_nh
)
```

```{r}
# Calculate 5-year average incarceration rates for each race
incarceration_3 <- incarceration_2 %>% 
  mutate(
         white_incarceration_rate_2015_2019 = 
           rowMeans(incarceration_2[,25:29], na.rm = TRUE), 
         
         black_incarceration_rate_2015_2019 = 
           rowMeans(incarceration_2[,30:34], na.rm = TRUE)
  )
```

```{r}
# Calculate RR for 2015-2019
incarceration_4 <- incarceration_3 %>% mutate(
  
  incarceration_2015_2019_RR = black_incarceration_rate_2015_2019 /
    white_incarceration_rate_2015_2019)
```

```{r}
# Create dataset for export
incarceration_4_select <- incarceration_4 %>% select(NAME, incarceration_2015_2019_RR)
colnames(incarceration_4_select) <- c("state", "incarceration-RR_2015-2019")

incarceration_for_export <- incarceration_4_select %>% pivot_longer(
  cols = contains("incarceration-RR"), 
  names_to = c("variable", "year"), 
  names_sep = "_", 
  values_to = "incarceration_RR"
)

incarceration_for_export %<>% mutate(year_renamed =
                                        case_when(
                                          grepl("2015-2019", year) ~ 2019) 
)

incarceration_for_export <- incarceration_for_export %>% select(state, year_renamed, incarceration_RR)

colnames(incarceration_for_export) <- c("state", "year", "incarceration_RR")

write_csv(incarceration_for_export, "../../data/cleaned/incarceration-nps-clean-for-viz.csv")

write_csv(incarceration_4, "../../data/cleaned/incarceration-nps-clean.csv")
```

```{r}
incarceration_look <- incarceration_4 %>% 
  pivot_longer(
  cols = starts_with("incarcerated"), 
  names_to = c("variable", "race", "ethnicity", "year"), 
  names_sep = "_", 
  values_to = "data")

incarceration_look2 <- incarceration_look %>% select(NAME, variable, race, year, data)

na.results <- incarceration_look2 %>% filter(is.na(data))

```
