---
title: "Homeownership data cleaning"
author: "Stephanie Veazie"
date: "2022-12-20"
output:
  word_document: default
  html_document: default
---

# Download data

1. Go to https://data.census.gov/cedsci/

2. Using the filter pane on the left-hand side of the page, under "Geography", select "State" then "All states with in the US, Puerto Rico, and the Island Areas." 

3. Using the search bar, search for "s2502" then select "S2502 Demographic characteristics for occupied housing units"

3. Under drop-down menu, select "2019: ACS 5-year estimates subject tables"

7. Select "transpose."

8. Download as a csv file. 

# Load libraries

```{r}
# Load libraries
library(tidyverse)
library(magrittr)
```

# Import files

```{r}
home_2019 <-read_csv("../../data/raw/us-census/ACSST5Y2019.S2502-2024-08-16T152131.csv")
```

# Clean 2019 data file

```{r}
# Create dataset of columns that report variable name and data by race
selected_homeownership <- home_2019 %>% 
  select(`Label (Grouping)`, 
         `Occupied housing units`, 
         `Occupied housing units!!RACE AND HISPANIC OR LATINO ORIGIN OF HOUSEHOLDER!!One race --!!Black or African American`,
         `Occupied housing units!!RACE AND HISPANIC OR LATINO ORIGIN OF HOUSEHOLDER!!White alone, not Hispanic or Latino`) 

names(selected_homeownership)

# Simplify the column names
names(selected_homeownership)<-c("variable", 
                       "total",
                       "black", 
                       "white_non_hispanic")

# Confirm change
names(selected_homeownership)

state_indices <- seq(1, 970, by = 19)

selected_homeownership %<>% 
  mutate(state_vector = case_when(row_number() %in% state_indices ~ variable,
                           !(row_number() %in% state_indices) ~ NA_character_)
  )

selected_homeownership %<>% mutate(grouping_indicator = ifelse(is.na(state_vector), 0, 1),
                                     group_sum = cumsum(grouping_indicator))

states_dat <- selected_homeownership %>% group_by(group_sum) %>% summarise(state_vector = first(state_vector))
selected_homeownership_data2 <- merge(states_dat, 
                                   selected_homeownership, 
                                   by = "group_sum",
                                   all.y = T)

selected_homeownership_data2 %<>%
  select(-state_vector.y, -starts_with("group")) %>% rename(state_vector = state_vector.x)

#select the total num occupied homes and num renter-occupied homes

selected_homeownership_data2 %<>% mutate(estimate_ind = ifelse(str_detect(variable, "Estimate"), 1, 0))

selected_homeownership_data2 %<>% group_by(state_vector) %>% mutate(est_count = cumsum(estimate_ind))

selected_homeownership_data3 <- selected_homeownership_data2 %>% filter(estimate_ind == 1 & est_count %in% c(1, 5)) %>% 
  mutate(variable = ifelse(estimate_ind == 1 & est_count == 1, "total_num_occupied_homes", "num_renter_occupied_homes"))

# Remove ',' and '%' from data in order to create numerical data
selected_homeownership_data4 <- selected_homeownership_data3 %>% mutate_all(~ str_replace(., "%", ""))
selected_homeownership_data4 <- selected_homeownership_data4 %>% mutate_all(~ str_replace_all(., ",", ""))

selected_homeownership_data4 %<>% select(-estimate_ind, -est_count)

# Pivot wider so each state's data is on 1 row.
tidy_homeownership <- selected_homeownership_data4 %>% 
  pivot_wider(names_from= variable, 
              values_from = c("total", "black", "white_non_hispanic"),
              names_glue="{variable}_{.value}")

#Confirm change
tidy_homeownership

tidy_homeownership %<>%  
  mutate(total_num_occupied_homes_black = as.numeric(total_num_occupied_homes_black),
         num_renter_occupied_homes_black = as.numeric(num_renter_occupied_homes_black),
         total_num_occupied_homes_white_non_hispanic = as.numeric(total_num_occupied_homes_white_non_hispanic),
         num_renter_occupied_homes_white_non_hispanic = as.numeric(num_renter_occupied_homes_white_non_hispanic))

# Create new dataset w/ RRs
tidy_homeownership %<>% 
  mutate(white_renter_occupied_homes_percent = 
           num_renter_occupied_homes_white_non_hispanic/
           total_num_occupied_homes_white_non_hispanic, 
         
         black_renter_occupied_homes_percent = 
           num_renter_occupied_homes_black / 
           total_num_occupied_homes_black
  ) 

tidy_homeownership %<>%  mutate(bw_renters_RR = black_renter_occupied_homes_percent / white_renter_occupied_homes_percent)

tidy_homeownership_2019 <- tidy_homeownership
```

```{r}
# Add year
homeownership_2019 <- tidy_homeownership_2019 %>% mutate(year = 2019)

homeownership_2019[, 2:10] <- lapply(homeownership_2019[,2:10], as.numeric)

# Rename state col
homeownership_2019 <- homeownership_2019 %>% rename(state = state_vector)

# Rearrange cols
homeownership_2019<- homeownership_2019[,c(1,11,2:10)]

# Export
write_csv(homeownership_2019, "../../data/cleaned/homeownership-clean.csv")
```

