---
title: "Calculate domain means, sd and B/W rates by typology"
author: "Stephanie Veazie"
date: "2022-12-20"
output:
  word_document: default
---

```{r}
# Load library
library(tidyverse)
library(foreign)
library(magrittr)
library(janitor)
```

# Import education data

```{r}
# Import data
no_college<- read_csv("../../data/cleaned/edu-attainment-clean.csv")

# Remove Puerto Rico
no_college <- no_college %>% filter(state != "Puerto Rico")

# Limit to 2 columns of interest
no_college_select <- no_college %>% select(state, year, percent_no_college_black, percent_no_college_white, edu_RR)

# 2019 dataset
no_college_select_2019 <- no_college_select %>% filter(year == 2019)
no_college_select_2019<- no_college_select_2019 %>% arrange(state)
```

# Import employment data

```{r}
# Import data
unemployment <- read_csv("../../data/cleaned/unemployment-clean.csv")

# Remove Puerto Rico
unemployment <- unemployment %>% filter(state != "Puerto Rico")

# Limit to 2 columns of interest
unemployment_select <- unemployment %>% select(state, year, unemployment_black, unemployment_white_non_hispanic, bw_unemploy_RR)

# 2019 dataset
unemployment_select_2019 <- unemployment_select %>% filter(year == 2019)
unemployment_select_2019<- unemployment_select_2019 %>% arrange(state)
```

# Import income

```{r}
# Import data
income <- read_csv("../../data/cleaned/income-clean.csv")

# Remove Puerto Rico
income <- income %>% filter(state != "Puerto Rico")

# Limit to 2 columns of interest
income_select <- income %>% select(state, year, median_income_black, median_income_white_non_hispanic, wb_income_RR)

# 2019 dataset
income_select_2019 <- income_select %>% filter(year == 2019)
income_select_2019<- income_select_2019 %>% arrange(state)
```

# Import homeownership data

```{r}
# Import data
homeownership <- read_csv("../../data/cleaned/homeownership-clean.csv")

# Remove Puerto Rico
homeownership <- homeownership %>% filter(state != "Puerto Rico")

# Limit to 2 columns of interest
homeownership_select <- homeownership %>% select(state, year, black_renter_occupied_homes_percent, white_renter_occupied_homes_percent, bw_renters_RR)

# Break into 2014 & 2019 datasets
homeownership_select_2019 <- homeownership_select %>% filter(year == 2019)
homeownership_select_2019<- homeownership_select_2019 %>% arrange(state)
```
# Import NCRP incarceration data

```{r}
# Import data
incarceration_ncrp <- read_csv("../../data/cleaned/incarceration-ncrp-clean.csv")

# rename state vector
incarceration_ncrp %<>% rename(state = state_vector)

# Limit to 3 columns of interest
incarceration_ncrp <- incarceration_ncrp %>% select(state, 
                                                    bw_rr_incarc, 
                                                    ave_incarceration_rate_per100k_black_nh, 
                                                    ave_incarceration_rate_per100k_white_nh)
```

# Import NPS incarceration data

```{r}
# Import data
incarceration_nps <- read_csv("../../data/cleaned/incarceration-nps-clean.csv")

# rename state fector
incarceration_nps %<>% rename(state = NAME)

# Limit to 3 columns of interest
incarceration_nps <- incarceration_nps %>% select(state,
                                                  white_incarceration_rate_2015_2019,
                                                  black_incarceration_rate_2015_2019,
                                                  incarceration_2015_2019_RR)
```

# Combine & export original data for 2019

```{r}
# Merge education & employment datasets
structural_racism_2019 <- left_join(no_college_select_2019, unemployment_select_2019, by = "state")
                              
# Add income dataset
structural_racism_2019 <- left_join(structural_racism_2019, income_select_2019, by = "state")

# Add renters dataset
structural_racism_2019 <- left_join(structural_racism_2019, homeownership_select_2019, by = "state")

# Add incarceration dataset
structural_racism_2019 <- left_join(structural_racism_2019, incarceration_ncrp, by = "state") 

# Remove cols
structural_racism_2019 %<>% select(-year.x, -year.y, -year.x.x, -year.y.y)

# Remove DC row
structural_racism_2019_no_dc <- structural_racism_2019 %>% filter(state != "District of Columbia")
```

# Swap out incarceration variable for sensitivity analysis results

```{r}
# Drop NCRP variables
structural_racism_2019_nps <- structural_racism_2019_no_dc %>% select(-c(ave_incarceration_rate_per100k_black_nh,
                                                                      ave_incarceration_rate_per100k_white_nh,
                                                                      bw_rr_incarc))
                                                                      
# Add NPS variable
structural_racism_2019_nps <- merge(structural_racism_2019_nps, incarceration_nps, by= "state")
```

# Import LPA results for original analysis

```{r}
# Import LPA results
lpa_results <- read_csv("../../data/cleaned/lpa_all_classes_2019_original_no_dc.csv")

#Recode data to switch 1 and 2 for three_class for visualization purposes 
lpa_results_2 <- lpa_results %>% mutate(three_class_reordered = 
                                          case_when(three_class == 2 ~ 1,
                                                    three_class == 1 ~ 2,
                                                    three_class == 3 ~ 3))
#Factor data
lpa_results_2$three_class_reordered<-as.factor(lpa_results_2$three_class_reordered)

# Save as dataframe
lpa_results_2<-as.data.frame(lpa_results_2)

lpa_results_3_class <- lpa_results_2 %>% select(state, 
                                                two_class, 
                                                three_class_reordered, 
                                                four_class, 
                                                five_class)
```

Import LPA results for NPS analysis 

```{r}
# Import LPA results
lpa_results_nps <- read_csv("../../data/cleaned/lpa_all_classes_2019_alternative_no_dc.csv")

#Recode data to switch 1 and 2 for three_class for visualization purposes 
lpa_results_nps <- lpa_results_nps %>% mutate(three_class_reordered = 
                                          case_when(three_class == 1 ~ 1,
                                                    three_class == 2 ~ 3,
                                                    three_class == 3 ~ 2))
# Save as dataframe
lpa_results_nps<-as.data.frame(lpa_results_nps)

lpa_results_nps <- lpa_results_nps %>% select(state, three_class_reordered)
```

```{r}
# Merge original SR data & class data
df <- merge(structural_racism_2019_no_dc, lpa_results_3_class, by.x = "state", by.y = "state")
df <- data.frame(df)
```

```{r}
# Merge SR with NPS data & class data
df_nps <- merge(structural_racism_2019_nps, lpa_results_nps, by = "state")
```

# Two class model with original NCRP data

```{r}
# create df of mean & sd of education RR by class  
edu_RR_df <- df %>%
  group_by(two_class) %>%
  summarize(edu_RR_mean = mean(edu_RR, na.rm = TRUE),
            edu_RR_sd = sd(edu_RR, na.rm = TRUE) ) %>%
  as.data.frame()

# create df of mean & sd of unemployment RR by class  
unemploy_RR_df<- df %>%
  group_by(two_class) %>%
  summarize(unemploy_RR_mean = mean(bw_unemploy_RR, na.rm = TRUE), 
            unemploy_RR_sd = sd(bw_unemploy_RR, na.rm = TRUE)) %>%
  as.data.frame()

# combine datasets
dataset <- left_join(edu_RR_df, unemploy_RR_df, by = "two_class")

# create df of mean & sd of renters RR by class  
renters_mean_df <- df %>%
  group_by(two_class) %>%
  summarize(renter_RR_mean = mean(bw_renters_RR, na.rm = TRUE), 
            renter_RR_sd = sd(bw_renters_RR, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, renters_mean_df, by = "two_class")

# create df of mean & sd of incarceration RR by class 
incarc_mean_df <-  df %>%
  group_by(two_class) %>%
  summarize(incarc_RR_mean = mean(bw_rr_incarc, na.rm = TRUE), 
            incarc_RR_sd = sd(bw_rr_incarc, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, incarc_mean_df, by = "two_class")

# create df of mean & sd of income RR by class  
med_income_df <- df %>%
  group_by(two_class) %>%
  summarize(income_RR_mean = mean(wb_income_RR, na.rm = TRUE), 
            income_RR_sd = sd(wb_income_RR, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, med_income_df, by = "two_class")

# convert factors to numeric
dataset$two_class <- as.numeric(dataset$two_class)
str(dataset)

# get table into correct format
dataset_t <- t(dataset) # transpose the table to get rownames as column headers
dataset_t <- dataset_t %>% row_to_names(row_number = 1) # change 1st row to header row
dataset_t <- round(dataset_t, 2) 

# formatting
plain_dataset <- rbind(
cbind(
paste0(dataset_t[1,1], " (", dataset_t[2,1], ")"),
paste0(dataset_t[1,2], " (", dataset_t[2,2], ")")),

cbind(
paste0(dataset_t[3,1], " (", dataset_t[4,1], ")"),
paste0(dataset_t[3,2], " (", dataset_t[4,2], ")")),

cbind(
paste0(dataset_t[5,1], " (", dataset_t[6,1], ")"),
paste0(dataset_t[5,2], " (", dataset_t[6,2], ")")),

cbind(
paste0(dataset_t[7,1], " (", dataset_t[8,1], ")"),
paste0(dataset_t[7,2], " (", dataset_t[8,2], ")")),

cbind(
paste0(dataset_t[9,1], " (", dataset_t[10,1], ")"),
paste0(dataset_t[9,2], " (", dataset_t[10,2], ")"))
)

rownames  <- c("Education inequity", 
               "Employment inequity",
               "Homeownership inequity", 
               "Incarceration inequity",
               "Income inequity"
               )

plain_dataset <- cbind(rownames, plain_dataset)

plain_dataset <- data.frame(plain_dataset)

colnames(plain_dataset) <- c("Domain", "Type 1", "Type 2")

plain_dataset

 write_csv(plain_dataset, "../../data/cleaned/typology_mean_sd_2_class.csv")
```

# Three class model with original NCRP data

```{r}
# create df of mean & sd of education RR by class  
edu_RR_df <- df %>%
  group_by(three_class_reordered) %>%
  summarize(edu_RR_mean = mean(edu_RR, na.rm = TRUE),
            edu_RR_sd = sd(edu_RR, na.rm = TRUE) ) %>%
  as.data.frame()

# create df of mean & sd of unemployment RR by class  
unemploy_RR_df<- df %>%
  group_by(three_class_reordered) %>%
  summarize(unemploy_RR_mean = mean(bw_unemploy_RR, na.rm = TRUE), 
            unemploy_RR_sd = sd(bw_unemploy_RR, na.rm = TRUE)) %>%
  as.data.frame()

# combine datasets
dataset <- left_join(edu_RR_df, unemploy_RR_df, by = "three_class_reordered")

# create df of mean & sd of renters RR by class  
renters_mean_df <- df %>%
  group_by(three_class_reordered) %>%
  summarize(renter_RR_mean = mean(bw_renters_RR, na.rm = TRUE), 
            renter_RR_sd = sd(bw_renters_RR, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, renters_mean_df, by = "three_class_reordered")

# create df of mean & sd of incarceration RR by class 
incarc_mean_df <-  df %>%
  group_by(three_class_reordered) %>%
  summarize(incarc_RR_mean = mean(bw_rr_incarc, na.rm = TRUE), 
            incarc_RR_sd = sd(bw_rr_incarc, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, incarc_mean_df, by = "three_class_reordered")

# create df of mean & sd of income RR by class  
med_income_df <- df %>%
  group_by(three_class_reordered) %>%
  summarize(income_RR_mean = mean(wb_income_RR, na.rm = TRUE), 
            income_RR_sd = sd(wb_income_RR, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, med_income_df, by = "three_class_reordered")

# convert factors to numeric
dataset$three_class_reordered <- as.numeric(dataset$three_class_reordered)
str(dataset)

# get table into correct format
dataset_t <- t(dataset) # transpose the table to get rownames as column headers
dataset_t <- dataset_t %>% row_to_names(row_number = 1) # change 1st row to header row
dataset_t <- round(dataset_t, 2) 

# formatting
plain_dataset <- rbind(
cbind(
paste0(dataset_t[1,1], " (", dataset_t[2,1], ")"),
paste0(dataset_t[1,2], " (", dataset_t[2,2], ")"),
paste0(dataset_t[1,3], " (", dataset_t[2,3], ")")),

cbind(
paste0(dataset_t[3,1], " (", dataset_t[4,1], ")"),
paste0(dataset_t[3,2], " (", dataset_t[4,2], ")"),
paste0(dataset_t[3,3], " (", dataset_t[4,3], ")")),

cbind(
paste0(dataset_t[5,1], " (", dataset_t[6,1], ")"),
paste0(dataset_t[5,2], " (", dataset_t[6,2], ")"),
paste0(dataset_t[5,3], " (", dataset_t[6,3], ")")),

cbind(
paste0(dataset_t[7,1], " (", dataset_t[8,1], ")"),
paste0(dataset_t[7,2], " (", dataset_t[8,2], ")"),
paste0(dataset_t[7,3], " (", dataset_t[8,3], ")")),

cbind(
paste0(dataset_t[9,1], " (", dataset_t[10,1], ")"),
paste0(dataset_t[9,2], " (", dataset_t[10,2], ")"),
paste0(dataset_t[9,3], " (", dataset_t[10,3], ")"))
)

rownames  <- c("Education inequity", 
               "Employment inequity",
               "Homeownership inequity", 
               "Incarceration inequity", 
               "Income inequity")

plain_dataset <- cbind(rownames, plain_dataset)

plain_dataset <- data.frame(plain_dataset)

colnames(plain_dataset) <- c("Domain", "Type 1", "Type 2", "Type 3")

plain_dataset

write_csv(plain_dataset, "../../data/cleaned/typology_mean_sd_3_class.csv")
```

# Four class model with original NCRP data

```{r}
# create df of mean & sd of education RR by class  
edu_RR_df <- df %>%
  group_by(four_class) %>%
  summarize(edu_RR_mean = mean(edu_RR, na.rm = TRUE),
            edu_RR_sd = sd(edu_RR, na.rm = TRUE) ) %>%
  as.data.frame()

# create df of mean & sd of unemployment RR by class  
unemploy_RR_df<- df %>%
  group_by(four_class) %>%
  summarize(unemploy_RR_mean = mean(bw_unemploy_RR, na.rm = TRUE), 
            unemploy_RR_sd = sd(bw_unemploy_RR, na.rm = TRUE)) %>%
  as.data.frame()

# combine datasets
dataset <- left_join(edu_RR_df, unemploy_RR_df, by = "four_class")

# create df of mean & sd of renters RR by class  
renters_mean_df <- df %>%
  group_by(four_class) %>%
  summarize(renter_RR_mean = mean(bw_renters_RR, na.rm = TRUE), 
            renter_RR_sd = sd(bw_renters_RR, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, renters_mean_df, by = "four_class")

# create df of mean & sd of incarceration RR by class 
incarc_mean_df <-  df %>%
  group_by(four_class) %>%
  summarize(incarc_RR_mean = mean(bw_rr_incarc, na.rm = TRUE), 
            incarc_RR_sd = sd(bw_rr_incarc, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, incarc_mean_df, by = "four_class")

# create df of mean & sd of income RR by class  
med_income_df <- df %>%
  group_by(four_class) %>%
  summarize(income_RR_mean = mean(wb_income_RR, na.rm = TRUE), 
            income_RR_sd = sd(wb_income_RR, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, med_income_df, by = "four_class")

# convert factors to numeric
dataset$four_class <- as.numeric(dataset$four_class)
str(dataset)

# get table into correct format
dataset_t <- t(dataset) # transpose the table to get rownames as column headers
dataset_t <- dataset_t %>% row_to_names(row_number = 1) # change 1st row to header row
dataset_t <- round(dataset_t, 2) 

# formatting
plain_dataset <- rbind(
cbind(
paste0(dataset_t[1,1], " (", dataset_t[2,1], ")"),
paste0(dataset_t[1,2], " (", dataset_t[2,2], ")"),
paste0(dataset_t[1,3], " (", dataset_t[2,3], ")"), 
paste0(dataset_t[1,4], " (", dataset_t[2,4], ")") 
),

cbind(
paste0(dataset_t[3,1], " (", dataset_t[4,1], ")"),
paste0(dataset_t[3,2], " (", dataset_t[4,2], ")"),
paste0(dataset_t[3,3], " (", dataset_t[4,3], ")"),
paste0(dataset_t[3,4], " (", dataset_t[4,4], ")") 
),

cbind(
paste0(dataset_t[5,1], " (", dataset_t[6,1], ")"),
paste0(dataset_t[5,2], " (", dataset_t[6,2], ")"),
paste0(dataset_t[5,3], " (", dataset_t[6,3], ")"), 
paste0(dataset_t[5,4], " (", dataset_t[6,4], ")") 
),

cbind(
paste0(dataset_t[7,1], " (", dataset_t[8,1], ")"),
paste0(dataset_t[7,2], " (", dataset_t[8,2], ")"),
paste0(dataset_t[7,3], " (", dataset_t[8,3], ")"),
paste0(dataset_t[7,4], " (", dataset_t[8,4], ")")
),

cbind(
paste0(dataset_t[9,1], " (", dataset_t[10,1], ")"),
paste0(dataset_t[9,2], " (", dataset_t[10,2], ")"),
paste0(dataset_t[9,3], " (", dataset_t[10,3], ")"),
paste0(dataset_t[9,4], " (", dataset_t[10,4], ")")
)
)

rownames  <- c("Education inequity", 
               "Employment inequity",
               "Homeownership inequity", 
               "Incarceration inequity",
               "Income inequity")

plain_dataset <- cbind(rownames, plain_dataset)

plain_dataset <- data.frame(plain_dataset)

colnames(plain_dataset) <- c("Domain", "Type 1", "Type 2", "Type 3", "Type 4")

plain_dataset

write_csv(plain_dataset, "../../data/cleaned/typology_mean_sd_4_class.csv")
```

# Five class model with original NCRP data

```{r}
# create df of mean & sd of education RR by class  
edu_RR_df <- df %>%
  group_by(five_class) %>%
  summarize(edu_RR_mean = mean(edu_RR, na.rm = TRUE),
            edu_RR_sd = sd(edu_RR, na.rm = TRUE) ) %>%
  as.data.frame()

# create df of mean & sd of unemployment RR by class  
unemploy_RR_df<- df %>%
  group_by(five_class) %>%
  summarize(unemploy_RR_mean = mean(bw_unemploy_RR, na.rm = TRUE), 
            unemploy_RR_sd = sd(bw_unemploy_RR, na.rm = TRUE)) %>%
  as.data.frame()

# combine datasets
dataset <- left_join(edu_RR_df, unemploy_RR_df, by = "five_class")

# create df of mean & sd of renters RR by class  
renters_mean_df <- df %>%
  group_by(five_class) %>%
  summarize(renter_RR_mean = mean(bw_renters_RR, na.rm = TRUE), 
            renter_RR_sd = sd(bw_renters_RR, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, renters_mean_df, by = "five_class")

# create df of mean & sd of incarceration RR by class 
incarc_mean_df <-  df %>%
  group_by(five_class) %>%
  summarize(incarc_RR_mean = mean(bw_rr_incarc, na.rm = TRUE), 
            incarc_RR_sd = sd(bw_rr_incarc, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, incarc_mean_df, by = "five_class")


# create df of mean & sd of income RR by class  
med_income_df <- df %>%
  group_by(five_class) %>%
  summarize(income_RR_mean = mean(wb_income_RR, na.rm = TRUE), 
            income_RR_sd = sd(wb_income_RR, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, med_income_df, by = "five_class")

# convert factors to numeric
dataset$five_class <- as.numeric(dataset$five_class)
str(dataset)

# get table into correct format
dataset_t <- t(dataset) # transpose the table to get rownames as column headers
dataset_t <- dataset_t %>% row_to_names(row_number = 1) # change 1st row to header row
dataset_t <- round(dataset_t, 2) 

# formatting
plain_dataset <- rbind(
cbind(
paste0(dataset_t[1,1], " (", dataset_t[2,1], ")"),
paste0(dataset_t[1,2], " (", dataset_t[2,2], ")"),
paste0(dataset_t[1,3], " (", dataset_t[2,3], ")"), 
paste0(dataset_t[1,4], " (", dataset_t[2,4], ")"), 
paste0(dataset_t[1,5], " (", dataset_t[2,5], ")") 
),

cbind(
paste0(dataset_t[3,1], " (", dataset_t[4,1], ")"),
paste0(dataset_t[3,2], " (", dataset_t[4,2], ")"),
paste0(dataset_t[3,3], " (", dataset_t[4,3], ")"),
paste0(dataset_t[3,4], " (", dataset_t[4,4], ")"), 
paste0(dataset_t[3,5], " (", dataset_t[4,5], ")")
),

cbind(
paste0(dataset_t[5,1], " (", dataset_t[6,1], ")"),
paste0(dataset_t[5,2], " (", dataset_t[6,2], ")"),
paste0(dataset_t[5,3], " (", dataset_t[6,3], ")"), 
paste0(dataset_t[5,4], " (", dataset_t[6,4], ")"), 
paste0(dataset_t[5,5], " (", dataset_t[6,5], ")") 
),

cbind(
paste0(dataset_t[7,1], " (", dataset_t[8,1], ")"),
paste0(dataset_t[7,2], " (", dataset_t[8,2], ")"),
paste0(dataset_t[7,3], " (", dataset_t[8,3], ")"),
paste0(dataset_t[7,4], " (", dataset_t[8,4], ")"), 
paste0(dataset_t[7,5], " (", dataset_t[8,5], ")")
),

cbind(
paste0(dataset_t[9,1], " (", dataset_t[10,1], ")"),
paste0(dataset_t[9,2], " (", dataset_t[10,2], ")"),
paste0(dataset_t[9,3], " (", dataset_t[10,3], ")"),
paste0(dataset_t[9,4], " (", dataset_t[10,4], ")"),
paste0(dataset_t[9,5], " (", dataset_t[10,5], ")")
)
)

rownames  <- c("Education inequity", 
               "Employment inequity",
               "Homeownership inequity", 
               "Incarceration inequity", 
               "Income inequity")

plain_dataset <- cbind(rownames, plain_dataset)

plain_dataset <- data.frame(plain_dataset)

colnames(plain_dataset) <- c("Domain", "Type 1", "Type 2", "Type 3", "Type 4", "Type 5")

plain_dataset

write_csv(plain_dataset, "../../data/cleaned/typology_mean_sd_5_class.csv")
```

# Three class model with NPS data (sensitivity analysis)

```{r}
# create df of mean & sd of education RR by class  
edu_RR_df <- df_nps %>%
  group_by(three_class_reordered) %>%
  summarize(edu_RR_mean = mean(edu_RR, na.rm = TRUE),
            edu_RR_sd = sd(edu_RR, na.rm = TRUE) ) %>%
  as.data.frame()

# create df of mean & sd of unemployment RR by class  
unemploy_RR_df<- df_nps %>%
  group_by(three_class_reordered) %>%
  summarize(unemploy_RR_mean = mean(bw_unemploy_RR, na.rm = TRUE), 
            unemploy_RR_sd = sd(bw_unemploy_RR, na.rm = TRUE)) %>%
  as.data.frame()

# combine datasets
dataset <- left_join(edu_RR_df, unemploy_RR_df, by = "three_class_reordered")

# create df of mean & sd of renters RR by class  
renters_mean_df <- df_nps %>%
  group_by(three_class_reordered) %>%
  summarize(renter_RR_mean = mean(bw_renters_RR, na.rm = TRUE), 
            renter_RR_sd = sd(bw_renters_RR, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, renters_mean_df, by = "three_class_reordered")

# create df of mean & sd of incarceration RR by class 
incarc_mean_df <-  df_nps %>%
  group_by(three_class_reordered) %>%
  summarize(incarc_RR_mean = mean(incarceration_2015_2019_RR, na.rm = TRUE), 
            incarc_RR_sd = sd(incarceration_2015_2019_RR, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, incarc_mean_df, by = "three_class_reordered")

# create df of mean & sd of income RR by class  
med_income_df <- df_nps %>%
  group_by(three_class_reordered) %>%
  summarize(income_RR_mean = mean(wb_income_RR, na.rm = TRUE), 
            income_RR_sd = sd(wb_income_RR, na.rm = TRUE)) %>%
  as.data.frame()

# add to dataset
dataset <- left_join(dataset, med_income_df, by = "three_class_reordered")

# convert factors to numeric
dataset$three_class_reordered <- as.numeric(dataset$three_class_reordered)
str(dataset)

# get table into correct format
dataset_t <- t(dataset) # transpose the table to get rownames as column headers
dataset_t <- dataset_t %>% row_to_names(row_number = 1) # change 1st row to header row
dataset_t <- round(dataset_t, 2) 

# formatting
plain_dataset <- rbind(
cbind(
paste0(dataset_t[1,1], " (", dataset_t[2,1], ")"),
paste0(dataset_t[1,2], " (", dataset_t[2,2], ")"),
paste0(dataset_t[1,3], " (", dataset_t[2,3], ")")),

cbind(
paste0(dataset_t[3,1], " (", dataset_t[4,1], ")"),
paste0(dataset_t[3,2], " (", dataset_t[4,2], ")"),
paste0(dataset_t[3,3], " (", dataset_t[4,3], ")")),

cbind(
paste0(dataset_t[5,1], " (", dataset_t[6,1], ")"),
paste0(dataset_t[5,2], " (", dataset_t[6,2], ")"),
paste0(dataset_t[5,3], " (", dataset_t[6,3], ")")),

cbind(
paste0(dataset_t[7,1], " (", dataset_t[8,1], ")"),
paste0(dataset_t[7,2], " (", dataset_t[8,2], ")"),
paste0(dataset_t[7,3], " (", dataset_t[8,3], ")")),

cbind(
paste0(dataset_t[9,1], " (", dataset_t[10,1], ")"),
paste0(dataset_t[9,2], " (", dataset_t[10,2], ")"),
paste0(dataset_t[9,3], " (", dataset_t[10,3], ")"))
)

rownames  <- c("Education inequity", 
               "Employment inequity",
               "Homeownership inequity", 
               "Incarceration inequity", 
               "Income inequity")

plain_dataset <- cbind(rownames, plain_dataset)

plain_dataset <- data.frame(plain_dataset)

colnames(plain_dataset) <- c("Domain", "Type 1", "Type 2", "Type 3")

plain_dataset

write_csv(plain_dataset, "../../data/cleaned/typology_mean_sd_3_class_nps.csv")
```


# Calculate B & W rates for each typology in 3 typology model 

```{r}
# Subset by class
class1<- df %>% filter(three_class_reordered == 1)
class2<- df %>% filter(three_class_reordered == 2)
class3<- df %>% filter(three_class_reordered == 3)
```

```{r}
# Education
class1_mean_edu_black <- mean(class1$percent_no_college_black)
class1_mean_edu_white <- mean(class1$percent_no_college_white)

class2_mean_edu_black <- mean(class2$percent_no_college_black)
class2_mean_edu_white <- mean(class2$percent_no_college_white)

class3_mean_edu_black <- mean(class3$percent_no_college_black)
class3_mean_edu_white <- mean(class3$percent_no_college_white)

# mean black & white rates by class    
class1_mean_edu_black 
class1_mean_edu_white

class2_mean_edu_black 
class2_mean_edu_white

class3_mean_edu_black
class3_mean_edu_white

# mean b-w ratio by class
class1_mean_edu_rr <- class1$edu_RR %>% mean(na.rm = TRUE)
class1_mean_edu_sd <- class1$edu_RR %>% sd(na.rm = TRUE)
class2_mean_edu_rr <- class2$edu_RR %>% mean(na.rm = TRUE)
class2_mean_edu_sd <- class2$edu_RR %>% sd(na.rm = TRUE)
class3_mean_edu_rr <- class3$edu_RR %>% mean(na.rm = TRUE)
class3_mean_edu_sd <- class3$edu_RR %>% sd(na.rm = TRUE)

class1_mean_edu_rr
class1_mean_edu_sd

class2_mean_edu_rr
class2_mean_edu_sd

class3_mean_edu_rr
class3_mean_edu_sd
```

```{r}
# Employment
class1_mean_unemploy_black <- mean(class1$unemployment_black)
class1_mean_unemploy_white <- mean(class1$unemployment_white_non_hispanic)

class2_mean_unemploy_black <- mean(class2$unemployment_black)
class2_mean_unemploy_white <- mean(class2$unemployment_white_non_hispanic)

class3_mean_unemploy_black <- mean(class3$unemployment_black)
class3_mean_unemploy_white <- mean(class3$unemployment_white_non_hispanic)

# mean b & w rates
class1_mean_unemploy_black
class1_mean_unemploy_white

class2_mean_unemploy_black
class2_mean_unemploy_white

class3_mean_unemploy_black
class3_mean_unemploy_white

# mean b-w ratio by class
class1_mean_unemploy_rr <- class1$bw_unemploy_RR %>% mean(na.rm = TRUE)
class1_mean_unemploy_sd <- class1$bw_unemploy_RR %>% sd(na.rm = TRUE)
class2_mean_unemploy_rr <- class2$bw_unemploy_RR %>% mean(na.rm = TRUE)
class2_mean_unemploy_sd <- class2$bw_unemploy_RR %>% sd(na.rm = TRUE)
class3_mean_unemploy_rr <- class3$bw_unemploy_RR %>% mean(na.rm = TRUE)
class3_mean_unemploy_sd <- class3$bw_unemploy_RR %>% sd(na.rm = TRUE)

class1_mean_unemploy_rr
class1_mean_unemploy_sd

class2_mean_unemploy_rr
class2_mean_unemploy_sd

class3_mean_unemploy_rr
class3_mean_unemploy_sd
```

```{r}
# Homeownership
class1_mean_renter_black <- mean(class1$black_renter_occupied_homes_percent)
class1_mean_renter_white <- mean(class1$white_renter_occupied_homes_percent)

class2_mean_renter_black <- mean(class2$black_renter_occupied_homes_percent)
class2_mean_renter_white <- mean(class2$white_renter_occupied_homes_percent)

class3_mean_renter_black <- mean(class3$black_renter_occupied_homes_percent)
class3_mean_renter_white <- mean(class3$white_renter_occupied_homes_percent)

# mean b & w rates
class1_mean_renter_black
class1_mean_renter_white

class2_mean_renter_black
class2_mean_renter_white

class3_mean_renter_black
class3_mean_renter_white

# mean b-w ratio by class
class1_mean_renter_rr <- class1$bw_renters_RR %>% mean(na.rm = TRUE)
class1_mean_renter_sd <- class1$bw_renters_RR %>% sd(na.rm = TRUE)
class2_mean_renter_rr <- class2$bw_renters_RR %>% mean(na.rm = TRUE)
class2_mean_renter_sd <- class2$bw_renters_RR %>% sd(na.rm = TRUE)
class3_mean_renter_rr <- class3$bw_renters_RR %>% mean(na.rm = TRUE)
class3_mean_renter_sd <- class3$bw_renters_RR %>% sd(na.rm = TRUE)

class1_mean_renter_rr
class1_mean_renter_sd

class2_mean_renter_rr
class2_mean_renter_sd

class3_mean_renter_rr
class3_mean_renter_sd
```

```{r}
# Incarceration
class1_mean_incarc_black <- mean(class1$ave_incarceration_rate_per100k_black_nh, na.rm= TRUE)
class1_mean_incarc_white <- mean(class1$ave_incarceration_rate_per100k_white_nh, na.rm= TRUE)

class2_mean_incarc_black <- mean(class2$ave_incarceration_rate_per100k_black_nh, na.rm= TRUE)
class2_mean_incarc_white <- mean(class2$ave_incarceration_rate_per100k_white_nh, na.rm= TRUE)

class3_mean_incarc_black <- mean(class3$ave_incarceration_rate_per100k_black_nh, na.rm= TRUE)
class3_mean_incarc_white <- mean(class3$ave_incarceration_rate_per100k_white_nh, na.rm= TRUE)

# B and W rates
class1_mean_incarc_black
class1_mean_incarc_white

class2_mean_incarc_black
class2_mean_incarc_white

class3_mean_incarc_black
class3_mean_incarc_white

# mean b-w ratio by class
class1_mean_incarc_rr <- class1$bw_rr_incarc %>% mean(na.rm = TRUE)
class1_mean_incarc_sd <- class1$bw_rr_incarc %>% sd(na.rm = TRUE)
class2_mean_incarc_rr <- class2$bw_rr_incarc %>% mean(na.rm = TRUE)
class2_mean_incarc_sd <- class2$bw_rr_incarc %>% sd(na.rm = TRUE)
class3_mean_incarc_rr <- class3$bw_rr_incarc %>% mean(na.rm = TRUE)
class3_mean_incarc_sd <- class3$bw_rr_incarc %>% sd(na.rm = TRUE)

class1_mean_incarc_rr
class1_mean_incarc_sd

class2_mean_incarc_rr
class2_mean_incarc_sd

class3_mean_incarc_rr
class3_mean_incarc_sd
```

```{r}
# Income
class1_mean_income_black <- mean(class1$median_income_black)
class1_mean_income_white <- mean(class1$median_income_white)

class2_mean_income_black <- mean(class2$median_income_black)
class2_mean_income_white <- mean(class2$median_income_white)

class3_mean_income_black <- mean(class3$median_income_black)
class3_mean_income_white <- mean(class3$median_income_white)

# B & W rates
class1_mean_income_black
class1_mean_income_white

class2_mean_income_black
class2_mean_income_white

class3_mean_income_black
class3_mean_income_white

# mean b-w ratio by class
class1_mean_income_rr <- class1$wb_income_RR %>% mean(na.rm = TRUE)
class1_mean_income_sd <- class1$wb_income_RR  %>% sd(na.rm = TRUE)
class2_mean_income_rr <- class2$wb_income_RR  %>% mean(na.rm = TRUE)
class2_mean_income_sd <- class2$wb_income_RR  %>% sd(na.rm = TRUE)
class3_mean_income_rr <- class3$wb_income_RR  %>% mean(na.rm = TRUE)
class3_mean_income_sd <- class3$wb_income_RR  %>% sd(na.rm = TRUE)

class1_mean_income_rr
class1_mean_income_sd

class2_mean_income_rr
class2_mean_income_sd

class3_mean_income_rr
class3_mean_income_sd
```

